---
title: "Components"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Components}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

A guided tour of the UI components in bslibdash. New here? Start with
`vignette("getting-started")` for an end-to-end app; for restyling see
`vignette("theming")`.

All examples assume:

```{r purl=FALSE}
library(shiny)
library(bslibdash)
```

## Cards

`box()` is the primary card. `boxLayout()` arranges several cards in a
responsive grid, and `updateBox()` controls a card from the server.

```{r purl=FALSE}
boxLayout(
  box("First card",  title = "Status",  status = "primary"),
  box("Second card", title = "Details", status = "info",
      collapsible = TRUE),
  type = "deck"
)

# server side
updateBox("status", action = "toggle")          # collapse/expand
updateBox("status", action = "update",
          options = list(title = "Updated", status = "success"))
```

Valid `status`/`background` values are the eight Bootstrap status names
(`primary`, `secondary`, `success`, `info`, `warning`, `danger`, `light`,
`dark`).

## Value & info boxes

```{r purl=FALSE}
fluidRow(
  valueBox("128", "Open tickets", icon = icon("inbox"),    color = "primary"),
  valueBox("42%", "CPU",          icon = icon("cpu"),      color = "warning"),
  infoBox(title = "Status",       value = "Healthy",       color = "success")
)
```

Both accept Bootstrap status names (`primary`, `success`, ...) **and**
the legacy `shinydashboard` palette (`aqua`, `green`, `red`, ...), which
is mapped automatically.

## Tab box

```{r purl=FALSE}
tabBox(
  id = "summary", title = "Q4",
  tabPanel("Overview", p("Overview content")),
  tabPanel("Details",  p("Detail content"))
)
```

## Sidebar navigation

```{r purl=FALSE}
dashboardSidebar(
  sidebarUserPanel("Jane Doe", subtitle = "Administrator"),
  sidebarSearchForm("q", "go", label = "Search records..."),
  sidebarMenu(
    id = "sidebarMenu",
    sidebarHeader("Main"),
    menuItem("Overview", tabName = "overview", icon = icon("house")),
    menuItem("Reports",  icon = icon("bar-chart"),
      menuSubItem("Daily",   tabName = "reports_daily"),
      menuSubItem("Monthly", tabName = "reports_monthly")
    )
  )
)
```

Navigate from the server with `updateTabItems()`:

```{r purl=FALSE}
observeEvent(input$go_reports, {
  updateTabItems(session, inputId = "sidebarMenu", selected = "reports")
})
```

## Header dropdown menus

```{r purl=FALSE}
dashboardHeader(
  title = "Operations",
  rightUi = dropdownMenu(
    type = "notifications", badgeStatus = "warning",
    notificationItem("Backup completed",  status = "success"),
    notificationItem("New deployment",    status = "info"),
    messageItem(from = "Ops bot", message = "Pipeline finished",
                color = "success"),
    taskItem(text = "Data refresh", value = 75, color = "info")
  )
)
```

Use `dropdownMenuOutput()` + `renderDropdownMenu()` to drive the panel
reactively. See the next section for the full set of output/render pairs.

## Reactive outputs and renderers

Most dynamic bslibdash components follow the familiar Shiny
`*Output()` + `render*()` pattern. Place the `*Output()` in the UI where
the component should appear, then build it from the server with the
matching `render*()`:

| UI side                   | Server side             | Re-renders                |
|---------------------------|-------------------------|---------------------------|
| `valueBoxOutput()`        | `renderValueBox()`      | A single `valueBox()`     |
| `infoBoxOutput()`         | `renderInfoBox()`       | A single `infoBox()`      |
| `dropdownMenuOutput()`    | `renderDropdownMenu()`  | A header `dropdownMenu()` |
| `sidebarMenuOutput()`     | `renderMenu()`          | A `sidebarMenu()`         |

## Accordion, badges, buttons, icons

```{r purl=FALSE}
accordion(
  id = "filters",
  accordionItem(title = "Date range", p("Last 30 days")),
  accordionItem(title = "Region",     p("All regions"), status = "info")
)

badge("NEW", color = "success", position = "right", rounded = TRUE)

actionButton("refresh", "Refresh", icon = "arrow-clockwise",
             status = "primary")

icon("user")                       # Bootstrap Icons via {bsicons}
icon("bar-chart", size = "1.25rem", class = "text-primary")
```

## Toasts

`toast()` is a thin wrapper over `shiny::showNotification()` that uses a
`bslib::card()` as the notification body:

```{r purl=FALSE}
observeEvent(input$save, {
  toast(
    title    = "Saved",
    body     = "Settings were updated.",
    options  = list(type = "message", delay = 3000),
    session  = session
  )
})
```
