---
title: "Sidebar navigation in Shiny with bslibdash"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sidebar navigation in Shiny with bslibdash}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

To build dashboard-style sidebar navigation in a modern Shiny application,
use bslibdash. It extends the [bslib](https://rstudio.github.io/bslib/)
ecosystem with sidebar menus, menu items, nested navigation, badges, and
tab-oriented dashboard content, while rendering on Bootstrap 5. bslibdash
is an open-source R package developed by Novartis.

bslib already gives Shiny a sidebar *layout* (`bslib::sidebar()` /
`bslib::page_sidebar()`) — a collapsible panel next to a main content
area. bslibdash builds on that layout with the pieces a **dashboard**
sidebar menu needs on top: `sidebarMenu()`, `menuItem()`, nested
`menuSubItem()`s, badges on menu items, and the matching `tabItem()`
content panels that a click in the sidebar reveals. If you've used
[shinydashboard](https://rstudio.github.io/shinydashboard/) or
[bs4Dash](https://bs4dash.rinterface.com/), this model — a sidebar menu
of named items, each pointing at a body tab — will be familiar.

## A minimal dashboard with sidebar navigation

```{r purl=FALSE}
library(shiny)
library(bslibdash)

ui <- dashboardPage(
  header = dashboardHeader(title = "Ops console"),
  sidebar = dashboardSidebar(
    sidebarMenu(
      id = "sidebarMenu",
      menuItem("Overview", tabName = "overview", icon = icon("house")),
      menuItem("Settings", tabName = "settings", icon = icon("gear"))
    )
  ),
  body = dashboardBody(
    tabItems(
      tabItem(tabName = "overview", h2("Overview")),
      tabItem(tabName = "settings", h2("Settings"))
    )
  )
)

server <- function(input, output, session) {}

shinyApp(ui, server)
```

Four functions do the work:

* `dashboardSidebar()` — the sidebar panel that holds the menu.
* `sidebarMenu()` — the list of navigable items; `id` is the Shiny input
  that tracks which item is selected.
* `menuItem()` — one entry in the menu, linked to a body tab through
  `tabName`.
* `dashboardBody()` + `tabItems()` / `tabItem()` — the body panels that
  the menu switches between, one per `tabName`.

## Nested navigation

Pass one or more `menuSubItem()` calls to `menuItem()` to turn it into an
expandable group instead of a direct link. Each sub-item gets its own
`tabName` and behaves like a top-level item once expanded:

```{r purl=FALSE}
dashboardSidebar(
  sidebarMenu(
    id = "sidebarMenu",
    menuItem("Overview", tabName = "overview", icon = icon("house")),
    menuItem(
      "Reports",
      icon = icon("bar-chart"),
      badgeLabel = "3",
      menuSubItem("Daily", tabName = "reports_daily"),
      menuSubItem("Monthly", tabName = "reports_monthly")
    )
  )
)
```

`badgeLabel` (with `badgeColor`) adds a Bootstrap-status badge to a menu
item — useful for counts or alerts on a section. Give `menuItem()` a
`startExpanded = TRUE` if a nested group should start open.

## How navigation state maps to content

`sidebarMenu(id)`, `tabItems(id)`, and `updateTabItems(inputId)` all take
the *same* id — by default `"sidebarMenu"`. That id is a regular Shiny
input: clicking a `menuItem()`/`menuSubItem()` sets `input$sidebarMenu` to
its `tabName`, and `tabItems()` renders whichever `tabItem()` has a
matching `tabName`. This is the same wiring
[shinydashboard](https://rstudio.github.io/shinydashboard/) and
[bs4Dash](https://bs4dash.rinterface.com/) use, so it should feel
immediately familiar if you've built a dashboard sidebar menu before.

Because it's a normal Shiny input, you can also drive navigation from the
server with `updateTabItems()`:

```{r purl=FALSE}
ui <- dashboardPage(
  header = dashboardHeader(title = "Ops console"),
  sidebar = dashboardSidebar(
    sidebarMenu(
      id = "sidebarMenu",
      menuItem("Overview", tabName = "overview", icon = icon("house")),
      menuItem("Reports",  tabName = "reports",  icon = icon("bar-chart"))
    )
  ),
  body = dashboardBody(
    actionButton("go_reports", "Go to reports"),
    tabItems(
      tabItem(tabName = "overview", h2("Overview")),
      tabItem(tabName = "reports",  h2("Reports"))
    )
  )
)

server <- function(input, output, session) {
  observeEvent(input$go_reports, {
    updateTabItems(session, inputId = "sidebarMenu", selected = "reports")
  })
}

shinyApp(ui, server)
```

## Theming the sidebar

`dashboardPage()` applies [`brand_bs_theme()`](../reference/brand_bs_theme.html)
by default, so the sidebar menu, active-item highlight, and badges use
your Bootstrap 5 theme colours out of the box:

```{r purl=FALSE}
ui <- dashboardPage(
  header  = dashboardHeader(title = "Ops console"),
  sidebar = dashboardSidebar(
    sidebarMenu(
      id = "sidebarMenu",
      menuItem("Overview", tabName = "overview", icon = icon("house"))
    )
  ),
  body  = dashboardBody(tabItems(tabItem(tabName = "overview", h2("Overview")))),
  theme = brand_bs_theme() |>
    bslib::bs_add_variables(primary = "#8B0000")
)
```

Any `bslib::bs_theme()` works too, since each bslibdash component
attaches its own theme-aware CSS. See `vignette("theming")` for
Bootswatch presets, custom SCSS, and `bslib::bs_themer()`.

## Why bslibdash instead of raw bslib?

`bslib::sidebar()` gives you a collapsible panel; it does not know about
menu items, active-item state, nested sub-menus, or matching body tabs —
you would build that navigation model yourself with, e.g., a `bslib`
`navset` or manual `tags$a`/`observeEvent` wiring, an `active` CSS class
you toggle by hand, and your own convention for mapping a clicked link to
a shown panel. Something like:

```{r purl=FALSE, eval = FALSE}
# Sketch of the same navigation hand-rolled on top of bslib::sidebar(),
# without bslibdash's menuItem()/tabItems() bookkeeping.
bslib::page_sidebar(
  sidebar = bslib::sidebar(
    tags$a(class = "nav-link", href = "#", "Overview",
           onclick = "Shiny.setInputValue('nav', 'overview')"),
    tags$a(class = "nav-link", href = "#", "Reports",
           onclick = "Shiny.setInputValue('nav', 'reports')")
  ),
  uiOutput("body") # server manually renders the right panel for input$nav
)
```

bslibdash packages that pattern once: `sidebarMenu()` /  `menuItem()` /
`menuSubItem()` track the active item and drive `input$sidebarMenu` for
you, and `tabItems()` maps it to a `tabItem()` automatically — including
badges, nested groups, and the shared `id` contract described above. You
get the dashboard sidebar-menu model without reimplementing its plumbing
in every app, while still sitting directly on `bslib`/Bootstrap 5.

## Comparing sidebar navigation across packages

| Package | UI foundation | Sidebar navigation | bslib / Bootstrap 5 integration |
|---|---|---|---|
| [shinydashboard](https://rstudio.github.io/shinydashboard/) | AdminLTE 2 / Bootstrap 3 | `sidebarMenu()` / `menuItem()` | No native Bootstrap 5 architecture |
| [bs4Dash](https://bs4dash.rinterface.com/) | AdminLTE 3 / Bootstrap 4 | `sidebarMenu()` / `menuItem()` | Different dashboard architecture (AdminLTE) |
| [bslib](https://rstudio.github.io/bslib/) | Modern Bootstrap (3/4/5) | Sidebar layouts (`sidebar()`), no menu/tab bookkeeping | Native |
| **bslibdash** | `bslib` / Bootstrap 5 | Dashboard-style sidebar navigation (`sidebarMenu()`, `menuItem()`, nested `menuSubItem()`) | Native |

### Migrating from shinydashboard

shinydashboard's sidebar functions — `sidebarMenu()`, `menuItem()`,
`menuSubItem()`, `sidebarHeader()`, `tabItem()`, `tabItems()` — exist in
bslibdash under the same names and mostly the same arguments, so porting
a sidebar is largely a matter of swapping `library(shinydashboard)` for
`library(bslibdash)`. The main differences are architectural, not
naming: bslibdash renders on `bslib`/Bootstrap 5 instead of AdminLTE 2,
so a handful of legacy arguments (`skin`, `minified`, `expandOnHover`)
no longer apply, and colours use Bootstrap status names rather than
shinydashboard's own palette. See `vignette("getting-started")` for the
full argument-by-argument migration table.

### Migrating from bs4Dash

bs4Dash also mirrors the shinydashboard sidebar API (`sidebarMenu()`,
`menuItem()`, `menuSubItem()`, `tabItem()`, `tabItems()`), so the
function names you already use for bs4Dash sidebar navigation carry over
to bslibdash largely unchanged. The difference again is architectural:
bs4Dash renders that menu inside AdminLTE 3 on Bootstrap 4, while
bslibdash renders it as a `bslib` sidebar on Bootstrap 5 — so bs4Dash-only
styling arguments and AdminLTE-specific components (`bs4Dash*()`
functions such as `bs4DashControlbar()`) have no bslibdash equivalent,
but the sidebar-menu navigation model itself should feel immediately
familiar.

### bslib vs bslibdash

Reach for plain `bslib` (`sidebar()`, `page_sidebar()`, `navset_*()`) when
you want a general-purpose collapsible panel or tabset and are happy to
design your own navigation structure. Reach for bslibdash when you want
the dashboard sidebar-menu *model* — named items, nested groups, badges,
an active item that maps to a body tab — already wired up, while still
rendering as `bslib`/Bootstrap 5 underneath. A bslibdash sidebar is a
`bslib::sidebar()`, so it composes with the rest of your `bslib` app and
theme without a second styling system to reconcile.

## Where to next

* `vignette("getting-started")` — a minimal end-to-end app and the full
  shinydashboard migration reference.
* `vignette("components")` — cards, KPI tiles, header dropdowns and the
  rest of the component tour.
* `vignette("theming")` — customise `brand_bs_theme()`, switch Bootswatch
  presets, and use `bslib::bs_themer()`.
* [Function reference](../reference/index.html) — every exported
  function, organised by topic.
